---
title: "STA310 HW3"
format: pdf
editor: visual
author: "Olivia Fu"
date: "2025-02-07"
---

```{r include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)
```

```{r}
library(ggplot2)
```

## Exercise 1

The probability mass function of a Poisson random variable is given by:

$$
P(X = x) = \frac{\lambda^x e^{-\lambda}}{x!}, \quad x = 0, 1, 2, \dots
$$

The likelihood function is:

$$
\begin{aligned}L(\lambda) &= \prod_{i=1}^n P(X_i = x_i)\\&= \prod_{i=1}^n \frac{\lambda^{x_i} e^{-\lambda}}{x_i!}\\&= \lambda^{\sum_{i=1}^n x_i} \cdot e^{-n\lambda} \cdot \prod_{i=1}^n \frac{1}{x_i!}\end{aligned}
$$

\clearpage

## Exercise 2

The log-likelihood function is:

$$
\begin{aligned}\ell(\lambda) &= \log L(\lambda) \\&=\log \left( \lambda^{\sum_{i=1}^n x_i}\cdot e^{-n\lambda} \cdot \prod_{i=1}^n \frac{1}{x_i!} \right)\\&= \log(\lambda^{\sum_{i=1}^n x_i}) + \log(e^{-n\lambda}) + \log(\prod_{i=1}^n \frac{1}{x_i!})\\&= \sum_{i=1}^n x_i \cdot \log \lambda - n\lambda + \sum_{i=1}^n \log \frac{1}{x_i!}\end{aligned}
$$

The first derivative of the log-likelihood function is:

$$
\frac{\partial \ell(\lambda)}{\partial \lambda} = \frac{\sum_{i=1}^n x_i}{\lambda} - n
$$

We then set the derivative equal to zero to find the MLE:

$$
\begin{aligned} \frac{\sum_{i=1}^n x_i}{\hat{\lambda}} - n = 0& \\\hat{\lambda} = \frac{\sum_{i=1}^n x_i}{n}&\end{aligned}
$$

The second derivative of the log-likelihood function is:

$$
\frac{\partial^2 \ell(\lambda)}{\partial \lambda^2} = -\frac{\sum_{i=1}^n x_i}{\lambda^2}
$$

Since $\lambda > 0$ and $x_i \geq 0$, the second derivative is negative (exclude the trivial case where all $x_i=0$), confirming that the critical point corresponds to a maximum.

\clearpage

## Exercise 3

```{r}
# Given data
n <- 100
sum_x <- 500

# Log-likelihood function for Poisson distribution
log_likelihood_function <- function(lambda) {
  if (lambda <= 0) {
    return(-Inf)
  } else {
    # The constant term can be ignored as 
    # it does not affect the maximization of the log-likelihood
    return(sum_x * log(lambda) - n * lambda)
  }
}

# Set up a sequence of lambda values
lambda_values <- seq(1, 10, length.out = 100000)

# Calculate the log-likelihood for each lambda
log_likelihood_values <- sapply(lambda_values, log_likelihood_function)

# Find the lambda that maximizes the log-likelihood
mle_lambda <- lambda_values[which.max(log_likelihood_values)]

# Print the MLE for lambda
cat("MLE for lambda:", mle_lambda, "\n")
```

```{r}
# Create a dataframe
data <- data.frame(lambda = lambda_values, log_likelihood = log_likelihood_values)

# Plot
ggplot(data, aes(x = lambda, y = log_likelihood)) +
  geom_line(color = "blue", size = 1) +
  geom_vline(xintercept = mle_lambda, color = "red", linetype = "dashed") +
  annotate("text", 
           x = mle_lambda + 0.6, 
           y = max(log_likelihood_values) - 40,
           label = paste("MLE =", round(mle_lambda, 3)), 
           color = "red") +
  labs(title = "Log-Likelihood for Poisson",
       x = "Lambda",
       y = "Log-Likelihood") +
  theme_minimal()
```

Therefore, the approximate MLE is 5, which matches the formula we derived in Exercise 2.

$$
\hat{\lambda} = \frac{\sum_{i=1}^n x_i}{n} = \frac{500}{100}=5
$$

\clearpage

## Exercise 4

### (a)

| **Game**  | **First five shots** | **Likelihood (no hot hand)** | **Likelihood (hot hand)**                                                                        |
|---------------|---------------|---------------|---------------------------|
| 1         | BMMBB                | $p_B^3(1 - p_B)^2$           | $(p_B)(1-p_{B\vert B})(1-p_B)(p_B)(p_{B\vert B}) = (p_B)^2(1-p_{B\vert B})(1-p_B)(p_{B\vert B})$ |
| 2         | MBMBM                | $p_B^2(1 - p_B)^3$           | $(1-p_B)(p_B)(1-p_{B\vert B})(p_B)(1-p_{B\vert B}) = (p_B)^2(1-p_{B\vert B})^2(1-p_B)$           |
| 3         | MMBBB                | $p_B^3(1 - p_B)^2$           | $(1-p_B)(1-p_B)(p_B)(p_{B\vert B})(p_{B\vert B}) = (p_B)(1-p_B)^2(p_{B\vert B})^2$               |
| 4         | BMMMB                | $p_B^2(1 - p_B)^3$           | $(p_B)(1-p_{B\vert B})(1-p_B)(1-p_B)(p_B) = (p_B)^2(1-p_{B\vert B})(1-p_B)^2$                    |
| 5         | MMMMM                | $(1 - p_B)^5$                | $(1-p_B)(1-p_B)(1-p_B)(1-p_B)(1-p_B) = (1-p_B)^5$                                                |
|           |                      |                              |                                                                                                  |
| **Total** |                      | $p_B^{10}(1 - p_B)^{15}$     | $(p_B)^7(1-p_B)^{11}(p_{B\vert B})^3(1-p_{B\vert B})^4$                                          |

### (b)

```{r}
likelihood <- function(pb){
  pb^(10) * (1 - pb)^(15)
}
likelihood(0.4)
likelihood(0.3)
```

When we substitute 0.4 and 0.3 for $p_B$​ in the likelihood function, we find that $p_B=0.4$ produces a higher likelihood than $p_B=0.3$ ($4.93 \times 10^{-8}$ vs. $2.80 \times 10^{-8}$). This means that the data is more consistent with $p_B=0.4$.

Furthermore, intuitively, if the data shows that the player made 10 baskets out of 25 attempts, it makes sense to estimate the probability of making a basket close to 0.4 (10/25), rather than lower values like 0.3, which are less aligned with the observed data.

### (c)

**MLE for No Hot Hand Model**

The likelihood and log-likelihood functions are:

$$
\begin{aligned}Lik(p_B) &= p_B^{10}(1 - p_B)^{15}\\\log(Lik(p_B)) &= \log(p_B^{10})+\log((1 - p_B)^{15})\\&= 10\log(p_B) + 15\log(1 - p_B)\end{aligned}
$$

We then take the first derivative of the log-likelihood function and set it to zero:

$$
\begin{aligned}\frac{d}{dp_B}log(Lik(p_B)) &= \frac{10}{p_B}-\frac{15}{1-p_B} = 0\\&\Rightarrow \frac{10}{p_B} = \frac{15}{1-p_B}\\&\Rightarrow 10(1-p_B)=15p_B\\&\Rightarrow 25p_B = 10\\&\hat{p_B} = \frac{10}{25} = 0.4\end{aligned}
$$

**MLE for Hot Hand Model**

The likelihood and log-likelihood functions are:

$$
$\begin{aligned}Lik(p_B, p_{B\vert B}) &= (p_B)^7(1-p_B)^{11}(p_{B\vert B})^3(1-p_{B\vert B})^4\\\log(Lik(p_B, p_{B\vert B})) &= \log((p_B)^7)+\log((1-p_B)^{11})+\log((p_{B\vert B})^3)+\log((1-p_{B\vert B})^4)\\&= 7\log((p_B))+11\log((1-p_B))+3\log((p_{B\vert B}))+4\log((1-p_{B\vert B}))\end{aligned}$
$$

We then take the first derivative of the log-likelihood function (with respect to $p_B$ and $p_{B \vert B}$ separately) and set them to zero:

$$
\begin{aligned}\frac{d}{dp_B}log(Lik(p_B, p_{B\vert B})) &= \frac{7}{p_B}-\frac{11}{1-p_B} = 0\\&\Rightarrow \frac{7}{p_B} = \frac{11}{1-p_B}\\&\Rightarrow 7(1-p_B)=11p_B\\&\Rightarrow 18p_B = 7\\&\hat{p_B} = \frac{7}{18} = 0.3889\\\\\frac{d}{dp_{B\vert B}}log(Lik(p_B, p_{B\vert B})) &= \frac{3}{p_{B\vert B}}-\frac{4}{1-p_{B\vert B}} = 0\\&\Rightarrow \frac{3}{p_{B\vert B}} = \frac{4}{1-p_{B\vert B}}\\&\Rightarrow 3(1-p_{B\vert B})=4p_{B\vert B}\\&\Rightarrow 7p_{B\vert B} = 3\\&\hat{p_{B \vert B}} = \frac{3}{7} = 0.4286\end{aligned}
$$

**Likelihood Ratio Test (LRT)**

No Hot Hand Model: $p_B$

Hot Hand Model: $p_B, p_{B \vert B}$

Hypothesis:

$H_0:p_B=p_{B \vert B}$

$H_0:p_B \neq p_{B \vert B}$

Firstly, we need to plug the MLEs into the log-likelihood function for each model to get the maximum value of the log-likelihood for each model.

```{r}
loglik1 <- function(pb){
  log(pb^(10) * (1 - pb)^(15))
}
loglik1(10/25)
```

```{r}
loglik2 <- function(pb, pbb) {
  log(pb^7 * (1-pb)^(11) * pbb^3 * (1-pbb)^4)
}
loglik2(7/18, 3/7)
```

Then, we use the Likelihood Ratio Test to determine if the difference is statistically significant

```{r}
LRT <- 2 * (loglik2(7/18, 3/7) - loglik1(10/25))
LRT
```

Therefore, LRT = 0.03292.

The test statistic follows a $\chi^2$ distribution with 1 degrees of freedom (the difference in the number of parameters between the two models). Therefore, the p-value is $P(\chi^2>LRT)$.

```{r}
pchisq(LRT, 1, lower.tail = FALSE)
```

Therefore, p-value = 0.8560.

The p-value is very larger, so we fail to reject $H_0$. We don't have convincing evidence that the hot hand model (conditional model) is an improvement over the no hot hand model (unconditional model). Thus, **there is no significant evidence that the hod hand exists**.
